Learning constraints in spreadsheets and tabular data

### 2016-09-08 Paper/ECAI/tabular.tex
applied the comments of Luc

\title{Learning \wat \sergey{here should be learning and not discovering, right?} constraints in spreadsheets and tabular data}
% see http://www.acm.org/binaries/content/assets/publications/article-templates/sig-alternate-sample.tex

@@

\begin{abstract}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, writing, maintaining and identifying good formulas for tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic discovery of constraints (formulas and relations) in raw tabular data in an unsupervised way.
Common spreadsheet formulas are represented as predicates whose arguments must satisfy the inherent properties of the function \wat \sergey{predicates but properties of a function? sounds fine?}. The challenge is to discover these without user information or training examples. % Knowledge discovery
We propose a two-stage generate and test method where the first stage uses constraint solving techniques to efficiently reduce candidates in a generic way based on the function signature. 
Our approach to this knowledge discovery \wat task takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
We show that we are able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

% why learn them
Such issues and mistakes could be overcome through the development of intelligent user-assisting tools. A core aspect of such tools would be the ability to automatically discover what constraints and formulae are present in a spreadsheet. This is the problem we study in this paper. We see multiple promising applications that can build on this, such as rediscovery constraints when importing raw data, auto-suggestion of formula's during spreadsheet use, automatic auto-completion of rows or columns, and consistency and error checking. In spirit, this builds on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from very few examples.

% why difficult
In this paper,  we investigate whether knowledge discovery \wat techniques can be used to learn constraints (formulas and other relations) in spreadsheet data in an unsupervised way.
From a machine learning point of view this is an unconventional problem because the data is in tabular form, but the constraints we wish to learn can involve both rows and columns of the table. Being unsupervised there is no labeled information either, although for every possible function one can verify whether a certain input satisfies the definition.
%Unlike in other knowledge discovery tasks, such as pattern mining~\cite{}, there are many different types of formula and constraints we wish to discover, each with their own signature (number and properties of arguments) and definition.
From a data mining point of view, the data is relational on the one hand, since we can have multiple tables with relationships between them, but also of mixed textual and numeric types. More closely related is clausal discovery~\cite{claudien,lallouet}, learning CSP constraints \cite{Quacq,Conacq,modelseeker} and dependency discovery in databases \cite{savnik}. But an important difference remains that we want to learn constraints on both columns and rows over integer, floating point as well as textual data.

% Moved up
The question that we ask in this paper is: is it possible to discover or reconstruct constraints in flat tabular spreadsheet data? The main challenge is the amount of combinations of rows and columns that need to be tried as input to the constraints.
To answer this challenge we propose a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
%
% contribs
Our contributions are as follows:
\begin{itemize}
\item we define the tabular constraint learning problem, where the goal is to learn constraints (functions and relations) that range over entire rows or columns in an unsupervised way;
\item we propose a 2-stage generate and test method that reasons for each candidate constraint over maximally contiguous blocks in the first stage, and only in the second stage looks at individual rows and columns and their content;
\item furthermore, in the first stage we use a constraint solver to efficiently enumerate all combinations of maximal blocks compatible with the candidate constraint's argument, using backtracking search and forward checking;
\item experiments on different publicly available spreadsheets show that the system is able to extract constraints with high precision and recall. 
\end{itemize}

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). Unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of a conjunction of literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. \sergey{that's where I added a new reference, needs to be checked and might need to be moved in the right place}. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here that we start not from existing formulas but from the data directly.

@@

\paragraph{Visual}
\sergey{Luc, can you check if the supplementary material at github is understandable for people? Do we have to add anything there like write that code would be available after?} The above assumptions do not hold for many tables. However, since the specification of tables is usually easy for humans, we propose a second approach which allows users to indicate tables using a visual tool (screenshots are available in the accompanying github repository with meta-data\footnote{\url{https://github.com/SergeyParamonov/TaCLe}\label{github-link}}).
Users select tables excluding the header and optionally specify an orientation.
The tool then generates the specification of the tables automatically.

\bibliographystyle{plain}
{
  \bibliography{references}}
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{journal_header}
\input{packages}
\input{macros}

\author{Name1 Surname1 \and Name2 Surname2 \and Name3 Surname3 \institute{----------------------} }

\begin{abstract}
  \sergey{Samuel! Add a reference to python-constraint somewhere, including the references list at the end}\\
  \sergey{Tias, Samuel, Luc, should we even mention miniZinc or ASP here, or is it not relevant?}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, modeling, maintaining and discovering formulas in tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic discovery of constraints (formulas and relations) in raw tabular data.
We see multiple promising applications for this technique, e.g. in rediscovering constraints, auto-completion and error checking.
Our method takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
Common spreadsheet functions are represented as predicates whose arguments must satisfy a number of constraints.
Constraint learning techniques are used to identify the predicates that hold between blocks of rows and columns of the tables.
We show that our approach is able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

\section{Introduction}
Millions of people across the world use spreadsheets every day.
The tabular representation of the data is often intuitive, and the programming of functions in individual cells is quickly learned.
However, large and complex sheets (possibly with multiple tables and relations between them) can be hard to handle.
Many end-users lack the understanding of the underlying structures and dependencies in such sheets and the data they contain.
This is especially the case when spreadsheets have been exported from other software such as Enterprise Resource Planning (ERP) systems.
In this case, often a comma-separated values (CSV) format is used meaning that all formulas are lost, including inter-sheet formulas and relations.
%This limited understanding is especially the case with spreadsheets exported from other software such as ERP packages. In this case,
Even in manually created spreadsheets, it can be challenging to be consistent and correct with formulas across big tables.
For example the influential Reinhart-Rogoff economical paper ``Growth in a Time of Debt'' had some of its claims contested~\cite{flaw_excel}, after an investigation of the used Excel sheets was shown to contain mistakes in formulae.

In this paper,  we investigate whether learning techniques can be used to infer constraints (formulas and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where header names such as \textbf{Total} and \textbf{Rank} already suggest the usage of spreadsheet operations such as \textit{SUM} or \textit{RANK}.
Looking at the first row in table~$T_2$ and the data in table~$T_1$, it is clear that the values in this row were computed by summing the values in the corresponding columns (\textbf{1st Quarter} through \textbf{Total}) of $T_1$.
Examining $T_1$ also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by summing the rows of the previous four columns.
This provides a flavor of how this problem is different from standard mining and learning settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple tables with relationships between them (e.g. foreign-keys). On the other hand, the data is most often mixed, e.g. textual and numeric. %, since people tend to use spreadsheets for financial and accounting computations.
Therefore, understanding and learning constraints in tabular data is a novel and challenging problem for machine learning.

% Moved up
The question that we answer in this paper is: is it possible to discover or reconstruct constraints in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
It operates directly on (headerless) tables of a spreadsheet in an unsupervised setting, that is, it reasons on raw tabular data directly with no example constraint instantiations given. %, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch additional application scenario's such as autocompletion and error detection.

The approach that we introduce borrows techniques from logical and relational learning \cite{luc_book}, where the discovery of clausal constraints has been studied in inductive logic programming \cite{claudien,lallouet}; and from constraint learning,  where several approaches to learning sets of constraints have been developed in a constraint programming setting \cite{Quacq,Conacq,modelseeker};
and from mining in databases, where inductive techniques have been applied to discover constraints (such as functional and multi-valued dependencies) in relational databases \cite{savnik}.  It also builds on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from
very few examples.
Our approach contrasts with these in that it focuses on learning both column- and row-constraints across tables on integer, floating point as well as textual data.

This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach and our problem statement. Section \ref{sec:approach} presents our approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents a more detailed discussion of the related work. Section \ref{sec:conclusions} provides conclusions.

% \textbf{Novelty:}
% \begin{itemize}
%   \item USED -- Unsupervised setting (contrary to flashfill, etc)
%   \item Numeric, different constraints (contrary to single textual function solution in flashfill, etc)
%   \item USED -- Data format (2D) -- data is no longer in rows like a classic ML or DM settings
%   \item USED -- Declarative, general / modular, stacking of constraint problems
% \end{itemize}
%}

@@

We focus on \textit{functional} constraints that could be used as formulas in spreadsheets, ignoring the structural constraints \textit{ALLDIFFERENT}, \textit{FOREIGNKEY} and \textit{PERMUTATION}.
Moreover, we included detected equalities as intended constraints (in 5~spreadsheets).
All experiments were run using Python 3.5.1 on a Macbook Pro, Intel Core i7 2.3~GHz with 16GB RAM.

@@

{\setlength{\tabcolsep}{0.33em}
\begin{table}[]
  \centering
  \caption{Summary of properties and results per spreadsheet category}
  \label{tbl:category_overview}
  \begin{tabularx}{\linewidth}{lccccccc}
 & \multicolumn{2}{c}{\textbf{Exercises} (9)} & \multicolumn{2}{c}{\textbf{Tutorials} (21)} & \multicolumn{2}{c}{\textbf{Data} (4)} \\
 & Overall & Sheet avg & Overall & Sheet avg & Overall & Sheet avg \\ \hline
    \textbf{Tables} & 19 & 2.11 & 48 & 2.29 & 4 & 1 \\ \hline
    \textbf{Cells} & 1231 & 137 & 1889 & 90 & 2320 & 580 \\ \hline
    \textbf{\begin{tabular}[c]{@{}l@{}}Intended\\[-4.5pt] Constraints\end{tabular}} & 34 & 3.78 & 52 & 2.48 & 6 & 1.50 \\ \hline \hline
\textbf{Recall} & 0.85 & 0.83 & 0.88 & 0.87 & 1.00 & 1.00 \\ \hline
\textbf{\begin{tabular}[c]{@{}l@{}}Recall\\[-4.5pt] Supported\end{tabular}} & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 \\ \hline
\textbf{Precision} & 0.97 & 0.98 & 0.70 & 0.91 & 1.00 & 1.00 \\ \hline
\textbf{Speed (s)} & 1.62 & 0.18 & 1.76 & 0.08 & 0.81 & 0.20
\end{tabularx}
\end{table}}

@@

\subsubsection*{Q2. How precise is \sname?}
Across all spreadsheets \sname achieves a precision of~$0.79~(81/102)$.
While the average precision per spreadsheet is over $0.90$ for all categories, the overall precision is much higher for the exercises and data categories ($0.97$ and $1.00$) than for the tutorials category ($0.70$).
Examining the tutorial spreadsheets confirms that there are a few spreadsheets that have a high number of additional constraints.
A qualitative analysis of these spreadsheets shows that their lower precision is caused by duplicate data or multiple ways to calculate the same vectors.
For these spreadsheets the precision could be increased by post-processing the constraints to detect implications or heuristically selecting one among multiple ways to compute the same vector.

@@

Of the 32 supported constraint templates (including equality), the slowest by far is fuzzy lookup using $\pm 36\%$ of the running time.
It is followed by product ($\pm 5\%$), relative difference ($\pm 5\%$) and various row-aggregates (each $3.5\% - 4\%$).

@@

\section{Related Work}\label{sec:related_work}
\sname combines ideas from several existing
approaches and different fields of research.

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). However, unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple “atomic” constraints rather than a set of rules that each consist of multiple literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy.
Also related to this line of research is the work on deriving constraints in databases such as functional and multi-valued dependencies \cite{savnik, Mannila-Raiha}
although that line of research has focused on more specialized techniques for specific types of constraints. Furthermore, the constraints and formulas used
in spreadsheets are often different from those in databases.

Second, there exist a number of algorithms in the constraint programming community that induce constraint (programs)
from one or more examples and questions. Two well-known approaches include ModelSeeker \cite{modelseeker} and Quacq \cite{Quacq}.
The former starts from a single example in the form of a vector of values and then exhaustively looks for the most specific constraints from a large constraint library that hold in the vector. To this aim, it cleverly enumerates different de-vectorized tables; for instance, if the initial vector was of dimension $1 \times 20$, it would consider rearrangements of size $2 \times 10$, $4 \times 5$, $5 \times 4$, … and Modelseeker would then look for both column, row  and even diagonal constraints. Key differences are that we assume given spreadsheets with \textit{multiple} tables; and that ModelSeeker's constraints are restricted to integer values and often typical for constraint programming community, e.g. aimed at combinatorial optimization and scheduling. Conacq \cite{Conacq} acquires a conjunction of constraints using techniques inspired on Mitchell's version spaces to process examples; its successor Quacq uses active learning to query the user regarding satisfiability. Each of the examples/queries is processed for each constraint, hence an important issue is how fast the algorithm converges to the entire solution set (e.g. for Sudoku).
The focus is again on integer variables and typical combinatorial optimisation constraints.  \sergey{that's where I added a new reference, needs to be checked and might need to be moved in the right place}. It also connects with the approach of inducing constraints and functions by generalizing given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here that we start not from existing formulas but from the data directly.
%As the conjunction for typical constraint programming problems (such as Sudoku) can be quite large, a key problem is to cope with redundant constraints. As Modelseeker, these systems focus on the types of constraints in combinatorial optimization but unlike ModelSeeker and our approach, they do not assume the variables are organized in tabular form.

\sergey{added database dependency discovery; have a look at it} Fourth, our work is motivated by the functional dependency discover in databases \cite{flach_dependency_discovery}. Most of the discovery techniques rely on the database schema that does not directly apply in our case. Some techniques work directly on the data \cite{tane_dependency_discovery} and, contrary to our work, are focused on pure functional dependencies that are not typical constraints of a spreadsheet that might hold over multiple tables and on a mixture of columns and rows.

@@

\bibliographystyle{plain}
{
  \bibliography{references}}
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

@@

\begin{abstract}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, writing, maintaining and identifying good formulas for tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic discovery of constraints (formulas and relations) in raw tabular data in an unsupervised way.
Common spreadsheet formulas are represented using predicates and expressions whose arguments must satisfy the inherent properties of the function. The challenge is to discover these without user information or labeled examples. 
We propose a two-stage generate and test method where the first stage uses constraint solving techniques to efficiently reduce candidates in a generic way based on the function signature. 
Our approach to this knowledge discovery task takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
We show that we are able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

% why learn them
Such issues and mistakes could be overcome through the development of intelligent user-assisting tools. A core aspect of such tools would be the ability to automatically discover what constraints and formulae are present in a spreadsheet. This is the problem we study in this paper. We discuss multiple promising applications that can build on this, such as rediscovery constraints when importing raw data, auto-suggestion of formula's during spreadsheet use, automatic auto-completion of rows or columns, and consistency and error checking. Our techniques build on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from very few examples.

% why difficult
In this paper,  we investigate whether machine learning and knowledge discovery techniques can be used to learn constraints (formulas and other relations) in spreadsheet data in an unsupervised way.
From a machine learning point of view this is an unconventional problem because the data is in tabular form, but the constraints we wish to learn can involve both rows and columns of the table. Being unsupervised there is no labeled information either, although for every possible function one can verify whether a certain input satisfies the definition.
%Unlike in other knowledge discovery tasks, such as pattern mining~\cite{}, there are many different types of formula and constraints we wish to discover, each with their own signature (number and properties of arguments) and definition.
From a data mining point of view, the data is relational on the one hand, since we can have multiple tables with relationships between them, but also of mixed textual and numeric types. More closely related is clausal discovery~\cite{claudien,lallouet}, learning CSP constraints \cite{Quacq,Conacq,modelseeker} and dependency discovery in databases \cite{savnik}. But an important difference remains that we want to learn constraints on both columns and rows over integer, floating point as well as textual data.

% Moved up
The question that we ask in this paper is: is it possible to discover or reconstruct constraints in flat tabular spreadsheet data? The main challenge is the number of combinations of rows and columns that need to be tried as input to the constraints.
To answer this challenge we propose a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
%
% contribs
Our contributions are as follows:
\begin{itemize}
\item we define the tabular constraint learning problem, where the goal is to learn constraints (functions and relations) that range over entire rows or columns in an unsupervised way;
\item we propose a 2-stage generate and test method that reasons for each candidate constraint over maximally contiguous blocks in the first stage, and only in the second stage looks at individual rows and columns and their content;
\item furthermore, in the first stage we use a constraint solver to efficiently enumerate all combinations of maximal blocks compatible with the candidate constraint's argument, using backtracking search and forward checking;
\item experiments on different publicly available spreadsheets show that the system is able to extract constraints with high precision and recall. 
\end{itemize}

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). However, unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of multiple literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. \sergey{that's where I added a new reference, needs to be checked and might need to be moved in the right place}. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here that we start not from existing formulas but from the data directly.

@@

@@

@@

@@

@@

@@

@@

\bibliographystyle{plain}
{
  \bibliography{references}}
\end{document}
